---
title: "Ophiocoma orientation modelling according to binomial distribution"
author: "John Kirwan"
date: "11/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load requisite packages.

```{r message=FALSE}
options(mc.cores = parallel::detectCores())
library('circular')
library('tibble')
library('readr')
library('tidyr')
library('dplyr')
library('ggplot2')
library('rstan')
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
rstan_options(auto_write = TRUE)
library('brms')
library(ggthemes)
theme_set(theme_tufte()) 
```

```{r}
sumlength <- function (x) { round(100* sum(x<0) / length(x),1) }
logistic  <- function (x) { p <- 1/(1 + exp(-x)); p <- ifelse(x == Inf, 1, p);   p }
logit     <- function (x) { log(x) - log(1 - x) }
```


The requisite part of pre-modelling markdown document is condensed below.

```{r message=FALSE, warning=FALSE}
df <- read_delim('Ophiocoma_orientation_data.txt',delim = "\t")
df <- subset.data.frame(df, Relative_heading_degrees != "DNF")
df$heading <- as.numeric(df$Relative_heading_degrees)*pi/180
df$heading[df$heading > pi] <- -pi +(df$heading[df$heading > pi] -pi)
df$Species <- as.factor(df$Species)
df$Type    <- as.factor(df$Type)
df$Set     <- as.factor(df$Set)
df <- subset.data.frame(df, select = -c(Absolute_heading_degrees,Relative_heading_degrees))
df$success   <- vector(length = length(df$heading))
sectors <- 5 # number of sectors, i.e. reciprocal of sector fraction
half.sector <- pi/sectors # half the width of the target sector
for(i in 1:length(df$heading)){           ### get tote of each using one fifth of the circle
  if(is.na(df$heading[i]) == TRUE){df$success[i] <- 0}
  else if(df$heading[i] >   half.sector ){df$success[i] <- 0}    ## pi/5 beforehand
  else if(df$heading[i] <= -half.sector ){df$success[i] <- 0}  ## -pi/5 beforehand
  else{df$success[i] <- 1}  }
df %>% filter(Set == 2017) -> df1
df1 <- droplevels.data.frame(df1)
df %>% filter(Set == 2019) -> df2
df2 <- droplevels.data.frame(df2)
```


# 2017 experiments with *Ophiocoma spp.*

The discrete visual (3 bars) stimulus in the 2017 set of experiments has been thus far included but, for simplicity, is excluded in the subsequent models, as the continuous visual (DoG) - the main effect - is used in preference. A differing discrete visual stimulus (the 3 bars variant) is, however, used for the 2019 experiments. The variant uses two white flanks of equal width of the dark target but half the amplitude. 


```{r}
Bin_Spp.formula <- bf(success ~ Species, family = bernoulli(link='logit'))
Bin_Spp.prior <- c(prior(normal(0, 2), class = "b"))

Bin_Spp.fit <- brm(Bin_Spp.formula,
               prior = Bin_Spp.prior, iter = 20000,
               data = df1, chains = 4,
               control = list(adapt_delta = 0.9))

Bin_Spp.fit <- add_criterion(Bin_Spp.fit, criterion=c('loo','waic'))

marginal_effects(Bin_Spp.fit, method = c("fitted"), 
                 resolution = 1000, robust=TRUE, 
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F) 
```

Posterior predictive check of counts.

```{r}
pp_check(Bin_Spp.fit, type="rootogram")
```


```{r}
Bin_Type.formula <- bf(success ~ Type, family = bernoulli(link='logit'))
Bin_Type.prior <- c(prior(normal(0, 2), class = "b"))

Bin_Type.fit <- brm(Bin_Type.formula,
               prior = Bin_Type.prior, iter = 20000,
               data = df1, chains = 4,
               control = list(adapt_delta = 0.9))

Bin_Type.fit <- add_criterion(Bin_Type.fit, criterion=c('loo','waic'))
```

The below plot indicates how orientation towards the stimulus differs for the three stimulus types, irrespective of the species. 

```{r}
marginal_effects(Bin_Type.fit, method = c("fitted"), resolution = 1000, robust=TRUE, 
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F) 
```

Posterior predictive check of counts.

```{r}
pp_check(Bin_Type.fit, type="rootogram")
```

## Both Species and stimulus type

```{r}
Bin_Both.formula <- bf(success ~ Species+Type, family = bernoulli(link='logit'))
Bin_Both.prior <- c(prior(normal(0, 2), class = "b"))

Bin_Both.fit <- brm(Bin_Both.formula,
               prior = Bin_Both.prior, iter = 20000,
               data = df1, chains = 4,
               control = list(adapt_delta = 0.9))

Bin_Both.fit <- add_criterion(Bin_Both.fit, criterion=c('loo','waic'))
```


The plot below indicates the effect of species on orientation towards the stimulus, when the species and stimulus type are modelled together but not the interaction between them. 

```{r}
marginal_effects(Bin_Both.fit, method = c("fitted"),
                 resolution = 1000, robust=TRUE, 
                 theme=theme_classic(), nsamples = 1000,
                                  re_formula = NULL,ask=F) 
```

Posterior predictive check of the count model, comparing expected counts to the data.

```{r}
pp_check(Bin_Both.fit, type="rootogram")
```


## Both Species and stimulus type with interaction

```{r}
Bin_Both_Interaction.formula <- bf(success ~ Species*Type, family = bernoulli(link='logit'))
Bin_Both_Interaction.prior <- c(prior(normal(0, 2), class = "b"))

Bin_Both_Interaction.fit <- brm(Bin_Both_Interaction.formula,
               prior = Bin_Both_Interaction.prior, iter = 20000,
               data = df1, chains = 4,
               control = list(adapt_delta = 0.99))

Bin_Both_Interaction.fit <- add_criterion(Bin_Both_Interaction.fit, criterion=c('loo','waic'))
summary(Bin_Both_Interaction.fit)
```


```{r}
marginal_effects(Bin_Both_Interaction.fit, method = c("fitted"), 
                 resolution = 1000, robust=TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F) 
```

Posterior predictive check of the count model, comparing expected counts to the data.

```{r}
pp_check(Bin_Both_Interaction.fit, type="rootogram")
```


Compare these discretized data models:

```{r}
loo_compare(Bin_Spp.fit,Bin_Type.fit,Bin_Both.fit,Bin_Both_Interaction.fit,criterion = 'waic')
```


```{r}
loo_compare(Bin_Spp.fit,Bin_Type.fit,Bin_Both.fit,Bin_Both_Interaction.fit,criterion=c("loo"))

```

The model for which only species is included is considered slightly more likely than those which varies only or stimulus type or both. The model where both effects are considered is slightly les likely, followed by that which also includes the interaction between them. These first three have very similar values (within the se range) and much more likely than the remaining model which considers stimulus type but not species. We continue with the model including both effects are their interaction.

```{r}
Bin_Both_Interaction.post <- posterior_samples(Bin_Both_Interaction.fit)
```


```{r message=FALSE}
pumila_1_bar     <- Bin_Both_Interaction.post$b_Intercept
pumila_3_bars    <- Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_Type3_bars
pumila_DoG       <- Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_TypeDoG
wendtii_1_bar    <- Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_Specieswendtii
wendtii_3_bars   <- Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_Specieswendtii + 
                     Bin_Both_Interaction.post$b_Type3_bars + Bin_Both_Interaction.post$`b_Specieswendtii:Type3_bars`
wendtii_DoG     <-  Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_Specieswendtii + 
                     Bin_Both_Interaction.post$b_TypeDoG + Bin_Both_Interaction.post$`b_Specieswendtii:TypeDoG`

vals <- as_tibble(cbind(pumila_1_bar,pumila_3_bars,pumila_DoG,wendtii_1_bar,wendtii_3_bars,wendtii_DoG))
vals <- gather(vals)

vals %>% mutate(value = logistic(value)) %>%
  ggplot(aes(key,value)) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  +
  theme_classic() + ylim(0,1) + labs(title = "Proportion") + ylab('proportions')
```


```{r Between Spp contrasts}
one_bar     <- wendtii_1_bar - pumila_1_bar
three_bars  <- wendtii_3_bars - pumila_3_bars
DoG         <- wendtii_DoG - pumila_DoG

Spp_contrasts <- as_tibble(cbind(one_bar,three_bars,DoG))
Spp_contrasts <- gather(Spp_contrasts)

Spp_contrasts %>%
  ggplot(aes(x=value,group=key,colour=key)) + 
  geom_density(size=1.5) + theme_classic() + geom_vline(xintercept = 0)

Spp_contrasts %>% gather %>% group_by(key) %>%
  summarise(Percentage = sumlength(value)) -> Overlap
Overlap
```

The above plot contrasts the estimates between the two species at each of the stimulus types to find the probability that they have the same orientedness. This is enumerated below as percentages in the table.

The Spp. clearly differ for the single dark bar (Phototactic) and main experimental DoG (continuous visual) stimuli but not so for the 3 bars (discrete visual) stimulus. 

The following code simulates random binomial data and then contrasts this with each of the treatments. 

```{r}
set.seed(555)
data <- rbinom(40000,40,1/sectors)/40
logit.data <- logit(data)

pumila_1_bar     <- Bin_Both_Interaction.post$b_Intercept - logit.data
pumila_3_bars    <- Bin_Both_Interaction.post$b_Intercept + 
                    Bin_Both_Interaction.post$b_Type3_bars - logit.data
pumila_DoG       <- Bin_Both_Interaction.post$b_Intercept + 
                    Bin_Both_Interaction.post$b_TypeDoG - logit.data
wendtii_1_bar    <- Bin_Both_Interaction.post$b_Intercept +     
                    Bin_Both_Interaction.post$b_Specieswendtii - logit.data
wendtii_3_bars   <- Bin_Both_Interaction.post$b_Intercept +
                    Bin_Both_Interaction.post$b_Specieswendtii + 
                    Bin_Both_Interaction.post$b_Type3_bars +
                    Bin_Both_Interaction.post$`b_Specieswendtii:Type3_bars` - logit.data
wendtii_DoG     <-  Bin_Both_Interaction.post$b_Intercept + Bin_Both_Interaction.post$b_Specieswendtii + 
                    Bin_Both_Interaction.post$b_TypeDoG +
                    Bin_Both_Interaction.post$`b_Specieswendtii:TypeDoG` - logit.data

control_contrasts <- as_tibble(cbind(
  pumila_1_bar,pumila_3_bars,pumila_DoG,wendtii_1_bar,wendtii_3_bars,wendtii_DoG))
control_contrasts <- gather(control_contrasts)

control_contrasts %>%
  ggplot(aes(x=value,group=key,colour=key)) + 
  geom_density(size=1.5) + theme_classic() + geom_vline(xintercept = 0)

control_contrasts %>% gather %>% group_by(key) %>%
  summarise(Percentage = sumlength(value)) -> Overlap
Overlap
```

These (above) are the contrast with the simulated random data (assuming 40 observations per treatment). The percentage can be interpreted as the probability that the data is at or below the value expected by random chance. 


This probability is small for the *O. wendtii* treatments using the bar (phototactic) and DoG (continuous visual) stimuli, indicating it is likely that they are oriented to the stimulus. It is larger for both *O. pumila* conditions, indicating it is plausible that they are disoriented. 


# 2019 experiments with *Ophiocoma wendtii* in differing diel and light-adaptation conditions


```{r}
Bin_Diel.formula <- bf(success ~ Period_adaptation, family = bernoulli(link='logit'))
Bin_Diel.prior <- c(prior(normal(0, 2), class = "b"))

Bin_Diel.fit <- brm(Bin_Diel.formula,
               prior = Bin_Diel.prior, iter = 20000,
               data = df2, chains = 4,
               control = list(adapt_delta = 0.9))

Bin_Diel.fit <- add_criterion(Bin_Diel.fit, criterion=c('loo','waic'))
```


The below plot indicates how orientation towards the stimulus differs for the four experimental conditions and the control.

```{r}
marginal_effects(Bin_Diel.fit, method = c("fitted"), resolution = 1000, robust=TRUE, 
                 theme=theme_classic(), nsamples = 1000, 
                                  re_formula = NULL,ask=F) 
```

```{r}
pp_check(Bin_Diel.fit, type="rootogram")
```

The marginal estimates at the median (of successful orientation for *O. wendtii* in differing diel and adaptive conditions) fall between 0.1 and 0.25, where the rate of success by random chance is 0.2, incidating the animals are disoriented. They are in the range of estimates for the earlier *O. pumila* treatments but not those of *O. wendtii* with the earlier bar (phototactic) and DoG (continuous visual) conditions.

The following contrasts the treatments with the control data. 

```{r}
Bin_Diel.post <- as_tibble(posterior_samples(Bin_Diel.fit))

Bin_Diel.post %>% select(-lp__,-b_Intercept) %>% gather %>%
  ggplot(aes(x=value,group=key,colour=key,fill=key,alpha=0.02)) + 
  geom_density() + theme_classic() + geom_vline(xintercept = 0)
```

These (below) are the contrast with the control data. The percentage can be interpreted as the probability that the data is at or below what would be expected by random chance. 

```{r}
Bin_Diel.post %>%  select(-lp__,-b_Intercept) %>% 
  gather %>% group_by(key) %>%
  summarise(Percentage = sumlength(value)) -> Overlap
Overlap
```


There is considerable overlap between these data and the control data and one treatment, the dark-adapted animals tested at night (Night_dark) is more uniform. 

We conclude given the estimates of successful orientation for each and their contrasts with the control that none of these treatments are oriented.



