---
title: "Ophiocoma von Mises modelling"
author: "John Kirwan"
date: "11/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Analyses to test whether observations of angular data are oriented towards stimuli, according to a von Mises distribution.

```{r message=FALSE}
options(mc.cores = parallel::detectCores())
library('circular')
library('tibble')
library('readr')
library('tidyr')
library('dplyr')
library('ggplot2')
library('rstan')
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
library('brms')
sumlength <- function (x) { round(100* sum(x<0) / length(x),1) }
```

The requisite part of pre-modelling markdown document is condensed below.

```{r message=FALSE, warning=FALSE}
df <- read_delim('Ophiocoma_orientation_data.txt',delim = "\t")
df <- subset.data.frame(df, Relative_heading_degrees != "DNF")
df$heading <- as.numeric(df$Relative_heading_degrees)*pi/180
df$heading[df$heading > pi] <- -pi +(df$heading[df$heading > pi] -pi)
df$Species <- as.factor(df$Species)
df$Type    <- as.factor(df$Type)
df$Set     <- as.factor(df$Set)
df <- subset.data.frame(df, select = -c(Absolute_heading_degrees,Relative_heading_degrees))
df %>% filter(Set == 2017) -> df2017
df2017 <- droplevels.data.frame(df2017)
df %>% filter(Set == 2019) -> df2019
df2019 <- droplevels.data.frame(df2019)
```

## "Null" model

To examine to what extent clustering occurs we can first run a model which looks at the observations but does not consider the species, giving a grand mean across the trials. The code below sets the model formula. 

```{r}
Mises_1_intercept.formula <- bf(heading ~ 0, kappa ~ 1, family = von_mises(link='tan_half'))
Mises_1_intercept.data    <- make_standata(heading ~ 1, kappa ~ 1,
                                  data = df2017, family = von_mises(link='tan_half'))
```

## Priors

To pick a suitable prior for the kappa value, we can look at different parameterizations of the gamma / lognormal distributions to use a positive bounded prior for kappa (concentration). The idea is to pick a reasonable range for the probability mass, without weighting it to heavily in either direction. The prior ideally ought to be *weakly informative*, giving little informationto influence the posterior, except to converge models and avoid nonsensical values. The priors are positive as they relate to kappa, rather than log(kappa).

```{r echo= FALSE}
plot(density((rgamma(20000,shape=2,scale=0.1))),xlim=c(-1,4),ylim=c(0,0.65),main="Prior choice")
lines(density((rlnorm(20000,0.25,0.75))),col="green")
lines(density((rlnorm(20000,0.1,1))),col="blue")
lines(density((rlnorm(20000,0,1))),col="red")
```

# Try replacing kappa intercept lognormal(0.1,1) later with something that doesn't penalise v low kappa, e.g. gamma(1,2)

The code below fits the model. The prior of lognormal(0.25, 0.75) is to guide it away from 0 to allow it converge.

```{r}
Mises_1_intercept.prior = c(prior(lognormal(0.25, 0.75),  class = Intercept, dpar="kappa"))

Mises_1_intercept.fit <- brm( Mises_1_intercept.formula,
  prior = Mises_1_intercept.prior, iter = 20000,
  data = df2017, chains = 4,
  control = list(adapt_delta = 0.9999)) 

Mises_1_intercept.fit <- add_criterion(Mises_1_intercept.fit, c("loo","waic")) 
summary(Mises_1_intercept.fit)
```

Rhats of 1 and an acceptable effective sample size in relation to the total number of obervations. The model converged well. Below is a posterior predictive plot.

```{r}
pp_check(Mises_1_intercept.fit, type="dens_overlay", nsamples=200)
```

Plot the posterior distributions and the Markov chains.

```{r echo=FALSE}
plot(Mises_1_intercept.fit,ask=F)
```

To interpret these, we can extract the values and reverse the log-link.

```{r echo=FALSE}
Mises_1_intercept.post <- posterior_samples(Mises_1_intercept.fit)
Mises_1_intercept.post$k_logged <- (Mises_1_intercept.post$b_kappa_Intercept)
Mises_1_intercept.post$R <- A1(exp(Mises_1_intercept.post$k_logged))
Mises_1_intercept.post %>% ggplot(aes(x=R)) + geom_density() + theme_classic()
```

And here is one with a dummy variable included to distinguish the signal types but still without species considered.

```{r echo=FALSE}
Mises_Type.formula <- bf(heading ~ 0, kappa ~ Type, family = von_mises(link='tan_half'))
Mises_Type.data    <- make_standata(heading ~ Type, kappa ~ 1,
                                  data = df2017, family = von_mises(link='tan_half'))

Mises_Type.prior = c(prior(normal(0,1),  class = Intercept, dpar="kappa"),
              prior(normal(0, 1), class = b, dpar="kappa")  )

Mises_Type.fit <- brm( Mises_Type.formula,
  prior = Mises_Type.prior, iter = 20000,
  data = df2017, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Type.fit <- add_criterion(Mises_Type.fit, c("loo","waic","R2")) 
```

Plot the posterior probabilities by parameter and the Markov chains.

```{r}
plot(Mises_Type.fit,ask=F)
```

Unimodal posterior distributions with well converged Markov chains.

```{r}
summary(Mises_Type.fit)
```

ESS and Rhat values are good. Below is a graphical check of the marginal effects at the median.

```{r}
marginal_effects(Mises_Type.fit, method = "fitted",dpar = "kappa", 
                 resolution = 1000, robust = TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F)
```

Posterior predictive check.

```{r}
pp_check(Mises_Type.fit, type="dens_overlay", nsamples=200)
```

We can implement a model which divvies it up by species.

```{r echo=FALSE}
Mises_Spp.formula <- bf(heading ~ 0, kappa ~ Species, family = von_mises(link='tan_half'))
Mises_Spp.data    <- make_standata(heading ~ Species, kappa ~ 1,
                                  data = df2017, family = von_mises(link='tan_half'))

Mises_Spp.prior = c(prior(normal(0, 1),  class = Intercept, dpar="kappa"),
              prior(normal(0, 1), class = b, dpar="kappa")  )

Mises_Spp.fit <- brm( Mises_Spp.formula,
  prior = Mises_Spp.prior, iter = 20000,
  data = df2017, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Spp.fit <- add_criterion(Mises_Spp.fit, c("loo","waic")) 
plot(Mises_Spp.fit,ask=F)
```

Summary of the von Mises model with the effect of species.

```{r}
summary(Mises_Spp.fit)
```

This converges well. Marginal effects at the median of the von Mises model with the effect of species.

```{r}
marginal_effects(Mises_Spp.fit,method = "fitted",dpar = "kappa", resolution = 1000, robust = TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F)
```

Posterior predictive check of the von Mises model with the effect of species.

```{r}
pp_check(Mises_Spp.fit, type="dens_overlay", nsamples=200)
```

Another version with the effects of Type and Species.

```{r}
Mises_Both.formula <- bf(heading ~ 0, kappa ~ Species + Type, family = von_mises(link='tan_half'))
Mises_Both.data    <- make_standata(heading ~ 1, kappa ~ Species*Type,
                                  data = df2017, family = von_mises(link='tan_half'))

Mises_Both.prior = c( prior(normal(0,1),  class = Intercept, dpar="kappa"),
               prior(normal(0, 1), class = b, dpar="kappa")
             )

Mises_Both.fit <- brm( Mises_Both.formula,
  prior = Mises_Both.prior, iter = 20000,
  data = df2017, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Both.fit <- add_criterion(Mises_Both.fit, criterion=c("loo","waic")) 
```

Model summary.

```{r}
summary(Mises_Both.fit)
```

```{r}
plot(Mises_Both.fit,ask=F)
```

Posterior predictive check.

```{r}
pp_check(Mises_Both.fit, type="dens_overlay", nsamples=200)
```

Marginal effect at the median.

```{r}
marginal_effects(Mises_Both.fit,method = "fitted",dpar = "kappa", resolution = 1000, robust = TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL,ask=F)
```


Further model with both effects and their interaction are included.

```{r}
Mises_Both_Interaction.formula <- bf(heading ~ 0, kappa ~ Species*Type, family = von_mises(link='tan_half'))
Mises_Both_Interaction.data    <- make_standata(heading ~ 1, kappa ~ Species*Type,
                                  data = df2017, family = von_mises(link='tan_half'))

Mises_Both_Interaction.prior = c( prior(normal(0,1),  class = Intercept, dpar="kappa"),
               prior(normal(0, 1), class = b, dpar="kappa"))

Mises_Both_Interaction.fit <- brm( Mises_Both_Interaction.formula,
  prior = Mises_Both_Interaction.prior, iter = 20000,
  data = df2017, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Both_Interaction.fit <- add_criterion(Mises_Both_Interaction.fit, criterion=c("loo","waic")) 
summary(Mises_Both_Interaction.fit)
```

```{r echo=FALSE}
plot(Mises_Both_Interaction.fit,ask=F)
```

```{r}
pp_check(Mises_Both_Interaction.fit, type="dens_overlay", nsamples=200)
```

```{r}
marginal_effects(Mises_Both_Interaction.fit, method = c("fitted"),
                 dpar = "kappa", resolution = 1000, robust = TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL, ask=F) 
```



Compare the loo (leave-one-out cross validation) values of the various models. 

```{r}
loo_compare(Mises_1_intercept.fit,Mises_Type.fit,Mises_Spp.fit,Mises_Both.fit,
            Mises_Both_Interaction.fit,criterion="loo")
```

Compare the waic (widely applicable information criterion) values of the various models. 

```{r}
loo_compare(Mises_1_intercept.fit,Mises_Type.fit,Mises_Spp.fit,Mises_Both.fit,
            Mises_Both_Interaction.fit,criterion="waic")
```

The von Mises model with the effect of species is the best fit - *O. wendtii* is more oriented than *O. pumila* regardless of the stimulus. 

## Full model with both effects and interaction

We interrogate with the model including both effects and their interaction and contrast the estimates of orientation with simulated data. 

```{r}
Mises_Both_Interaction.post <- posterior_samples(Mises_Both_Interaction.fit)
Mises_Both_Interaction.post %>%
      mutate( pumila_1_bar = ((b_kappa_Intercept)),
            pumila_DoG = ((b_kappa_Intercept + b_kappa_TypeDoG)),
            pumila_3_bars   = ((b_kappa_Intercept + b_kappa_Type3_bars)),
            wendtii_1_bar   = ((b_kappa_Intercept + b_kappa_Specieswendtii)),
            wendtii_DoG  = (
              (b_kappa_Intercept + b_kappa_Specieswendtii + 
                 `b_kappa_Specieswendtii:TypeDoG`)),
            wendtii_3_bars  = 
            ((b_kappa_Intercept + b_kappa_Specieswendtii + `b_kappa_Specieswendtii:Type3_bars`))) %>%
            dplyr::select(-c(lp__,b_kappa_Intercept,b_kappa_TypeDoG,b_kappa_Type3_bars,b_kappa_Specieswendtii,`b_kappa_Specieswendtii:TypeDoG`,`b_kappa_Specieswendtii:Type3_bars`)) -> Mises_Both_Interaction.effects 

Mises_Both_Interaction.effects %>% gather -> Mises_Both_Interaction.effects # df of exponentiated posteriors

Mises_Both_Interaction.effects %>%  mutate(value = A1(exp(value))) -> Mises_Both_Interaction.R   

Mises_Both_Interaction.effects %>%
  ggplot(aes(x=value,group=key,color=key,fill=key,alpha=0.1)) + geom_density() + theme_classic() + 
  geom_vline(xintercept=0) + xlab('log kappa')

```

Now, we plot the mean resultant length values.

```{r}
Mises_Both_Interaction.R %>%
  ggplot(aes(x=value,group=key,color=key,fill=key,alpha=0.1)) + geom_density() + theme_classic() +
  xlab('Mean resultant length')
```


```{r message=FALSE}
library(ggthemes)
theme_set(theme_tufte()) 

Mises_Both_Interaction.effects %>% 
  ggplot(aes(key,value)) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  +
  theme_classic() + labs(title = "kappa logged") 

Mises_Both_Interaction.effects %>% mutate(value = exp(value)) %>%
  ggplot(aes(key,value)) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  +
  theme_classic() + ylim(0,2.5) + labs(title = "kappa")

Mises_Both_Interaction.effects %>%
  ggplot(aes(key,A1(exp(value)))) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  + theme_classic() + labs(title = "Mean resultant length")
```

The first of the above plots shows the kappa values and the latter has these transformed to measures of R. In both cases it can be seen that *wendtii* is unlikely to be uniform, whereas *O. pumila* is uniform and unlikely to be very clustered. In both cases, I have also included the posterior distributions for the sum of the data, as before. These are more peaked than the results split by species because they are less uncertain, as there is more data (120 vs 40).

Exactly zero, in both cases, is theoretically uniform but what we expect from a random sample of 40 observations? We can check with some simple simulation.

```{r}
test1 <- rvon_mises(40,0,0)
test1 <- tibble(heading = test1)
test1.formula <- bf(heading ~ 0, kappa ~ 1, family = von_mises(link='tan_half') )
test1.prior = c(prior(normal(0, 3),  class = Intercept, dpar="kappa"))

test1.fit <- brm( test1.formula, prior = test1.prior, iter = 20000,
  data = test1, chains = 4, control = list(adapt_delta = 0.99999)) 

test1.fit <- add_criterion(test1.fit, c("loo","waic")) 
summary(test1.fit)
log.kappa.simulations <- posterior_samples(test1.fit)
log.kappa.simulations <- log.kappa.simulations$b_kappa_Intercept
```


Contrasts between the conditions and simulated random data.

```{r}
Mises_Both_Interaction.post %>%
  mutate( pumila_1_bar = b_kappa_Intercept - log.kappa.simulations,
    pumila_DoG     = b_kappa_Intercept + b_kappa_TypeDoG - log.kappa.simulations,
    pumila_3_bars  = b_kappa_Intercept + b_kappa_Type3_bars - log.kappa.simulations,
    wendtii_1_bar  = b_kappa_Intercept + b_kappa_Specieswendtii - log.kappa.simulations,
    wendtii_DoG    = b_kappa_Intercept + b_kappa_Specieswendtii + 
           `b_kappa_Specieswendtii:TypeDoG` - log.kappa.simulations,
            wendtii_3_bars = b_kappa_Intercept + b_kappa_Specieswendtii + 
            `b_kappa_Specieswendtii:Type3_bars` - log.kappa.simulations) %>%
    dplyr::select(-c(lp__,b_kappa_Intercept,b_kappa_TypeDoG,b_kappa_Type3_bars,
                     b_kappa_Specieswendtii,`b_kappa_Specieswendtii:TypeDoG`,
                     `b_kappa_Specieswendtii:Type3_bars`)) %>% gather -> Mises_Both_Interaction.contrasts 

Mises_Both_Interaction.contrasts  %>%
  ggplot(aes(x=value,group=key,color=key)) + geom_density(size=1.2) + theme_classic() + 
  geom_vline(xintercept=0) + xlab('log(kappa)') + xlim(-3,7)

Mises_Both_Interaction.contrasts %>% group_by(key) %>% 
      summarise(Proportion = sumlength(value)) -> Overlap
Overlap
```

Percentage overlap with simulated random data(above).

For all stimulus types, *O. wendtii* is highly unlikely to have the same orientedness or lower than the random simulated data - whereas, *O. pumila* has a much greater probability of being disoriented for all three stimulus types, especially the continuous visual stimulus (DoG).


# 2019 experiments with *Ophiocoma wendtii* in differing diel and light-adaptation conditions

The code below fits the model. 

```{r}
Mises_Diel_1_intercept.formula <- bf(
  heading ~ 0, kappa ~ 1, family = von_mises(link='tan_half'))

Mises_Diel_1_intercept.prior = c(prior(normal(0,1),  class = Intercept, dpar="kappa"))

Mises_Diel_1_intercept.fit <- brm( Mises_Diel_1_intercept.formula,
  prior = Mises_Diel_1_intercept.prior, iter = 20000,
  data = df2019, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Diel_1_intercept.fit <- add_criterion(Mises_Diel_1_intercept.fit, c("loo","waic"))
```

Rhats of 1 and a large effective sample size in relation to the total number of obervations. Plot the Markov chains and posterior porbabilities. 

```{r echo=FALSE}
plot(Mises_Diel_1_intercept.fit,ask=FALSE)
```


```{r}
pp_check(Mises_Diel_1_intercept.fit, type="dens_overlay", nsamples=200)
```

To interpret these, we can extract the values and reverse the log-link.

```{r echo=FALSE}
Mises_Diel_1_intercept.post <- posterior_samples(Mises_Diel_1_intercept.fit)
Mises_Diel_1_intercept.post$k_logged <- exp(Mises_Diel_1_intercept.post$b_kappa_Intercept)
Mises_Diel_1_intercept.post$R <- A1(exp(Mises_Diel_1_intercept.post$k_logged))

Mises_Diel_1_intercept.post %>%
  ggplot(aes(x=k_logged)) + geom_density() + theme_classic()

Mises_Diel_1_intercept.post %>%
  ggplot(aes(x=R)) + geom_density() + theme_classic()

```

And here is one with a dummy variable included to distingiush the signal types but still without species considered.

```{r echo=FALSE}
Mises_Diel_Period_adaptation.formula <- bf(heading ~ 0, kappa ~ Period_adaptation, family = von_mises(link='tan_half'))
Mises_Diel_Period_adaptation.data    <- make_standata(heading ~ Period_adaptation, kappa ~ 1,
          data = df2019, family = von_mises(link='tan_half'))

Mises_Diel_Period_adaptation.prior = c(prior(normal(0,1),  class = Intercept, dpar="kappa"),
              prior(normal(0, 1), class = b, dpar="kappa")  )

Mises_Diel_Period_adaptation.fit <- brm( Mises_Diel_Period_adaptation.formula,
  prior = Mises_Diel_Period_adaptation.prior, iter = 20000,
  data = df2019, chains = 4,
  control = list(adapt_delta = 0.9)) 

Mises_Diel_Period_adaptation.fit <- add_criterion(Mises_Diel_Period_adaptation.fit, c("loo","waic"))
summary(Mises_Diel_Period_adaptation.fit)
```

Plot the Markov chains and posterior probabilities. 

```{r}
plot(Mises_Diel_Period_adaptation.fit,ask=F)
```

Marginal effects at the median.

```{r}
marginal_effects(Mises_Diel_Period_adaptation.fit,method = "fitted",dpar = "kappa", resolution = 1000, robust = TRUE,
                 theme=theme_classic(), nsamples = 1000, scale='response',
                                  re_formula = NULL)
```

The night_dark treatment has a lowe kappa estimate as there is not orientation towards the stimulus. 

Posterior predictive checks.

```{r}
pp_check(Mises_Diel_Period_adaptation.fit, type="dens_overlay", nsamples=200)
```

This mean resultant length is a bit different from that calculated using circular. Now, we can implement a model which divvies it up by species.

```{r}
Mises_Diel_Period_adaptation.post <- posterior_samples(Mises_Diel_Period_adaptation.fit)
Mises_Diel_Period_adaptation.post %>%
      mutate( control = ((b_kappa_Intercept)),
    day_dark    =  ((b_kappa_Intercept + b_kappa_Period_adaptationday_dark)),
    night_dark   = ((b_kappa_Intercept + b_kappa_Period_adaptationnight_dark)),
    night_light   = ((b_kappa_Intercept + b_kappa_Period_adaptationnight_light)))  %>%
            dplyr::select(-c(
              lp__,b_kappa_Intercept,b_kappa_Period_adaptationday_dark,
              b_kappa_Period_adaptationnight_dark,
              b_kappa_Period_adaptationnight_light)) -> Mises_Diel_Period_adaptation.effects 

Mises_Diel_Period_adaptation.effects %>% gather -> Mises_Diel_Period_adaptation.effects 

Mises_Diel_Period_adaptation.effects %>%  mutate(value = A1(exp(value))) -> Mises_Diel_Period_adaptation.R   

Mises_Diel_Period_adaptation.effects %>%
  ggplot(aes(x=value,group=key,color=key,fill=key,alpha=0.1)) + geom_density() + theme_classic() + geom_vline(xintercept=0) 

Mises_Diel_Period_adaptation.R %>%
  ggplot(aes(x=value,group=key,color=key,fill=key,alpha=0.1)) + geom_density() + theme_classic() 
```



```{r message=FALSE}
library(ggthemes)
theme_set(theme_tufte()) 

Mises_Diel_Period_adaptation.effects %>% 
  ggplot(aes(key,value)) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  +
  theme_classic() + labs(title = "kappa logged") 

Mises_Diel_Period_adaptation.effects %>% mutate(value = exp(value)) %>%
  ggplot(aes(key,value)) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  +
  theme_classic() + ylim(0.5,2) + labs(title = "kappa")

Mises_Diel_Period_adaptation.effects %>%
  ggplot(aes(key,A1(exp(value)))) + geom_violin(aes(color=key,fill=key,alpha=0.1)) + geom_tufteboxplot()  + theme_classic() + labs(title = "Mean resultant length") + ylab('Mean resultant length')
```

Using these estimates of kappa (calcuated using MLE) as a baseline, we contrast the other measures against this.

The above densities are the difference between the posterior probabilities of kappa (concentration) for each Period_adaptation condition and the simulated 'random' uniform data.


# Comparing control dataset to experimental conditions

Then get contrasts for Period_adaptation types by unlogging the kappa values and finding the difference between them and the control to get a new distribution. The proportion overlap in the posterior density with zero indicates the probability that the two Period_adaptation conditons have the same value.


```{r Contrasts}
Mises_Diel_Period_adaptation.post %>%
    mutate(
    day_dark    =  b_kappa_Period_adaptationday_dark,
    night_dark   = b_kappa_Period_adaptationnight_dark,
    night_light   = b_kappa_Period_adaptationnight_light)  %>%
            dplyr::select(-c(
              lp__,b_kappa_Intercept,b_kappa_Period_adaptationday_dark,
              b_kappa_Period_adaptationnight_dark,
              b_kappa_Period_adaptationnight_light)) %>% gather -> Mises_Diel_Period_adaptation.contrasts 

Mises_Diel_Period_adaptation.contrasts %>%
  ggplot(aes(x=value,group=key,color=key)) + geom_density(size=1.2) + theme_classic() + 
  geom_vline(xintercept=0) + xlab('log(kappa)') + xlim(-4,4)
```

```{r}
Mises_Diel_Period_adaptation.contrasts %>% group_by(key) %>%
  summarise(Proportion = sumlength(value)) -> Overlap
Overlap
```

None of the experimental treatments have a low probability of having the same circular concentration as the control. 

